---
title: "Statistical inference with the GSS data"
author: "Dale Richardson"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
    keep_md: true
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

The `gss.Rdata` file is provided in my git-hub repository. 

```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

Observations in the GSS sample were collected using a form of random sampling (either `block quota sampling` in the years 1972-1974 and for half of the 1975-1976 surveys or `full probability` sampling for half of the 1975-1976 surveys and up to 2014.) Complete details of the sampling procedures are provided in Appendix A of the [documentation section of the GSS website](http://gss.norc.org/Get-Documentation).  

To put things more simply: "Each survey from 1972 to 2004 was an independently drawn sample of English-speaking persons 18 years of age or over, living in non-institutional arrangements within the United States. Starting in 2006 Spanish-speakers were added to the target population."

Therefore, the target population of the survey is English and Spanish speakers 18 years or older, residing in the United States in non-institutional arrangments, e.g. military installations, religious institutions, prison, etc. Since the observations were obtained using a random sampling approach, we can generalize our findings based on this sample to the entire United States adult population who speak English or Spanish.

Data were collected by in-person interviews either with paper questionnaires or starting in 2002, the GSS was conducted by using computer-assisted personal interviewing. Respondents to the GSS were randomly selected, but there was no random assignment of respondents to particular groups, meaning: this is an observational study, **NOT** an experiment, thus it is not possible to infer causality. 


* * *

## Part 2: Research question

For this short project, my research question is a simple one and considers the cumuluative responses across all years of the survey:

**Does gun ownership vary according to political ideology or is gun ownership independent of political ideology?**

This question is interesting to me because I am simply curious to see how gun ownership might vary according to political view. I assume that liberals would own less guns than conservatives, but I am fully aware that there are probably many other factors at play in determining whether someone owns a gun or not. Does the person live in a rural area or the city? Is an extremely conservative less likely to own a gun in the city, whereas an extremely liberal person more likely to own a gun if living in a rural area? I think I am interested in my proposed research question because it is a good springboard for further exploration into other demographic factors and variables that influence whether or not someone owns a gun.


* * *

## Part 3: Exploratory data analysis

Here, I will illustrate the steps I've taken to extract the necessary variables for exploratory analyses. I will focus on the political views variable, `polviews` and the gun-ownership variable, `owngun`. 

Recall that I am including responses across all years of the GSS, which means I have included responses from 1972-2012. 

```{r, data wrangling, fig.width = 10}

# select variables of interest and remove NAs 
poli.guns <- gss %>% select(polviews, owngun) %>% na.omit() 

#%>% filter(owngun != "Refused") %>% group_by(polviews, owngun) %>% summarise(counts = n()) %>% print(n=40)

# can use the table function in base R to tally results into a contigency table
poli.guns.tab <- table(poli.guns)

# filter out the "Refused" respondents by simply indexing the columns we want to keep
poli.guns.tab <- poli.guns.tab[,c(1,2)]

# print the table
poli.guns.tab

# Make a plot showing the distribution of counts
poli.guns.plot <- gss %>% select(polviews, owngun) %>% na.omit() %>% filter(owngun != "Refused") %>% group_by(polviews, owngun) 

ggplot(poli.guns.plot, aes(polviews, ..count..)) +
        geom_bar(aes(fill = owngun), position = "dodge") +
        theme_bw()
```

According to the above plot, we can immediately see that the ratio of gun owners to non-gun owners approaches 1 as one moves up the categories from "extremely liberal" to "extremely conservative". Another interesting facet of this data is that there are many more survey respondents that identify themselves are moderates. I would venture to say that the majority of Americans are probably moderates in their political views, while the extreme liberals and conservatives are well, relegated to the tail ends of a normal distribution (**NOT** depicted above!)

I am curious to see how the above plot would look if we turned it into a relative frequency bar plot. 

```{r rel-frequency bar plot, fig.width = 10}

# plot rel freq
ggplot(poli.guns.plot, aes(polviews)) +
        geom_bar(aes(fill = owngun), position = "fill") +
        theme_bw()

```

And as I suspected, this relative frequency bar plot clearly shows the shift in gun ownership as one becomes more conservative in ideology. One can see that for the "Conservative" and "Extremely Conservative" categories there are nearly 50% gun owners and non-gun owners (thus approaching a ratio of 1). 

* * *

## Part 4: Inference

```{r inference}

# Perform chi-square test of independence

chisq.test(poli.guns.tab) 

```

