---
title: "Statistical inference with the GSS data"
author: "Dale Richardson"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
    keep_md: true
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `gss`. Delete this note when before you submit 
your work. 

```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

How were the observations in the sample collected? Answers should discuss if random sampling was used. 

"Each survey from 1972 to 2004 was an independently drawn sample of English-speaking persons 18 years of age or over, living in non-institutional arrangements within the United States. 

Starting in 2006 Spanish-speakers were added to the target population. 

Block quota sampling was used in 1972, 1973, and 1974 surveys and for half of the 1975 and 1976 surveys. 

Full probability sampling was employed in half of the 1975 and 1976 surveys and the 1977, 1978, 1980, 1982-1991, 1993-1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012, and 2014 surveys. 

Also, the 2004, 2006, 2008, 2010, 2012, and 2014 surveys had sub-sampled non-respondents (see Appendix A for a detailed description of the sample design)."


What are the implications of this collection method on the scope of inference (generalizability/causality)? For causality, answer should discuss whether random assignment was used.

* * *

## Part 2: Research question



* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

* * *

## Part 4: Inference

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.