---
title: "Statistical inference with the GSS data"
author: "Dale Richardson"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
    keep_md: true
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

The `gss.Rdata` file is provided in my git-hub repository. 

```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

Observations in the GSS sample were collected using a form of random sampling (either `block quota sampling` in the years 1972-1974 and for half of the 1975-1976 surveys or `full probability` sampling for half of the 1975-1976 surveys and up to 2014.) Full details of the sampling procedure are provided in Appendix A of the [documentation section of the GSS website](http://gss.norc.org/Get-Documentation). 

Or, more simply: "Each survey from 1972 to 2004 was an independently drawn sample of English-speaking persons 18 years of age or over, living in non-institutional arrangements within the United States. Starting in 2006 Spanish-speakers were added to the target population."

Therefore, the target population of the survey is English and Spanish speakers 18 years or older, residing in the United States in non-institutional arrangments, e.g. military installations, religious institutions, prison, etc. Since the observations were obtained using a random sampling approach, we can generalize our findings based on this sample to the entire United States adult population who speak English or Spanish.

Data were collected by in-person interviews either with paper questionnaires or starting in 2002, the GSS was conducted by using computer-assisted personal interviewing. Respondents to the GSS were randomly selected, but there was no random assignment of respondents to particular groups, meaning: this is an observational study, **NOT** an experiment, thus it is not possible to infer causality. 


* * *

## Part 2: Research question



* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

* * *

## Part 4: Inference

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.